The state learner -- a super learner for right-censored data

### 2025-05-28 manuscript/manuscript-one.tex
typos

Accurate risk prediction from time-to-event data is a
important problem in many areas of medical research and public
health. Super learning \citep{van2007super}, also known as
ensemble learning or stacked regression
\citep{wolpert1992stacked,breiman1996stacked}, provides a
powerful approach to this problem by combining multiple
candidate prediction models. In the context of survival
analysis, a super learner can for instance combine Cox
regression models and random survival forests \citep[][Section
8.4]{gerds2021medical}.

In this paper we propose the state learner, a new super
learner designed to handle the specific challenges of ensemble
learning with right-censored data. The state learner estimates
the performance of a library of candidate models for both the
event-time and censoring distributions simultaneously. This
joint estimation framework avoids the need for pre-specifying
a censoring model, which is a limitation of many existing
methods. Our approach also allows for full flexibility in the
choice of models in the library, contrasting with earlier
proposals that restrict the library to specific model classes
\citep{polley2011-sl-cens,golmakani2020super}. Furthermore,
unlike iterative approaches such as those in
\cite{han2021inverse} and \cite{westling2021inference}, we can
provide theoretical guarantees for the performance of the
state learner.

@@

Super learning requires a library of candidate models, or
\textit{learners}, and a loss function for evaluating
performance on hold-out samples. Machine learning based on
right-censored data commonly uses the partial log-likelihood
as a loss function
\citep[e.g.,][]{li2016regularized,yao2017deep,lee2018deephit,katzman2018deepsurv,gensheimer2019scalable,lee2021boosted,kvamme2021continuous}.
This loss function is also suggested for super learning with
right-censored data by \cite{polley2011-sl-cens}, where data
is assumed to be observed in discrete time. However, the
partial log-likelihood loss does not work well as a measure of
performance in hold-out samples observed in continuous time.
The reason is that the partial log-likelihood loss assigns an
infinite value for any learner that predicts piece-wise
constant cumulative hazard functions when there are event
times in the test set which do not occur in the learning set.
This problem occurs with prominent survival learners including
the Kaplan-Meier estimator, random survival forests, and
semi-parametric Cox regression models and these learners
cannot be included in the library of the super learner
proposed by \cite{polley2011-sl-cens}. When a proportional
hazards model is assumed, the baseline hazard function can be
profiled out of the likelihood \citep{cox1972regression}. The
cross-validated partial log-likelihood loss
\citep{verweij1993cross} has therefore been suggested as a
loss function for super learning by \cite{golmakani2020super}.
This choice of loss function restricts the library of learners
to include only Cox proportional hazards models, and hence
excludes many learners such as, e.g., random survival forests,
additive hazards models, and accelerated failure time models.

@@

More recent work by \cite{han2021inverse} and
\cite{westling2021inference} circumvents the need to
pre-specify a censoring model by iterating
between estimation of the outcome and censoring models. However, it seems that this procedure is in general not
guaranteed to converge to the true data-generating mechanism
\citep[][Appendix~A.4]{munch2024thesis}.

@@

We introduce our notation and framework in
Section~\ref{sec:framework}. Section~\ref{sec:super-learning}
defines general super learning for right-censored data. In
Section~\ref{sec:super-learner-simple} we define the state
learner, while Section~\ref{sec:theor-results-prop} provides
theoretical guarantees. In Section~\ref{sec:use-cases-state}
we demonstrate how the state learner can be used to obtain
risk predictions and how it can be applied in the context of
causal inference. Section~\ref{sec:numer-exper} reports the
results of numerical experiments, and
Section~\ref{sec:real-data-appl} illustrates the method on
prostate cancer data. We conclude with a discussion in
Section~\ref{sec:discussion}. Proofs are collected in
Appendix~\ref{sec:proofs}. Code and an implementation of the
state learner are available at
\url{https://github.com/amnudn/statelearner}.

@@

\begin{proposition}
  \label{prop:stric-prop}
  If \(P \in\mathcal{P}\) then
  \begin{equation*}
    F_P = \argmin_{F \in \mathcal{H}_{\mathcal{P}}} P{[\bar{B}_\tau(F, \blank)]},
  \end{equation*}
  \( P \)-almost surely.
\end{proposition}
\begin{proof}
  See Appendix~\ref{sec:consistency-proof}.
\end{proof}

@@

Proposition~\ref{prop:oracle-prop} provides a precise bound on
the average price we pay for using cross-validation to select
a learner from a library of candidates as compared to knowing
the oracle learner in advance. Corollary~\ref{cor:asymp-cons}
states that this price is asymptotically vanishing as long as
the number of learners in the library grows with sample size
at a polynomial rate.

@@

\subsection{Targeted and debiased machine learning}
\label{sec:targeted-learning}

The framework of \textit{targeted} or \textit{debiased}
machine learning is a general methodology for combining
data-adaptive modeling, such as machine and super learning,
with asymptotically valid statistical inference for
interpretable target parameters \citep{van2006targeted,
  chernozhukov2018double}. \fxnote*{add more references?}{This
  framework has been applied extensively for causal inference
  based on observational data
  \citep[e.g.,][]{van2011targeted,kennedy2016semiparametric,rytgaard2022targeted}.}
A targeted or debiased estimator is constructed by first
estimation high-dimensional nuisance parameters using machine
learning, and then employing a debiasing or targeting step by
using tools from semi-parametric efficiency theory
\citep{pfanzagl1985contributions,bickel1993efficient,van2003unified,tsiatis2007semiparametric,kennedy2022semiparametric}.
In the context of right-censored competing risk data, the
high-dimensional nuisance parameters needed to be estimated
will typically include the cause-specific cumulative hazard
functions and the cumulative hazard function for censoring
\citep[e.g.,][]{van2003unified,rytgaard2021estimation,rytgaard2022targeted}.
The state learner is hence particularly well suited for
targeted and debiased machine learning as it immediately
provides estimates of these nuisance parameters. Though it
served as our original motivation for developing the state
learner, we leave the study of the state learner in the
context of targeted and debiased machine learning for a future
paper.

>>>>

Accurate risk prediction from time-to-event data is an
important problem in many areas of medical research and public
health. Super learning \citep{van2007super}, also known as
ensemble learning or stacked regression
\citep{wolpert1992stacked,breiman1996stacked}, provides a
powerful approach to this problem by combining multiple
candidate prediction models. In the context of survival
analysis, a super learner can for instance combine Cox
regression models and random survival forests \citep[][Section
8.4]{gerds2021medical}.

In this paper we propose the state learner, a new super
learner designed to handle the specific challenges of ensemble
learning with right-censored data. The state learner estimates
the performance of a library of candidate models for both the
event-time and censoring distributions simultaneously. This
joint estimation framework avoids the need for pre-specifying
a censoring model, which is a limitation of many existing
methods. Our approach also allows for full flexibility in the
choice of models in the library, contrasting with other
proposals that restrict the library to specific model classes
\citep{polley2011-sl-cens,golmakani2020super}. Furthermore,
unlike iterative approaches such as those in
\cite{han2021inverse} and \cite{westling2021inference}, we can
provide theoretical guarantees for the performance of the
state learner.

@@

Super learning requires a library of candidate models, or
\textit{learners}, and a loss function for evaluating
performance on hold-out samples. Machine learning based on
right-censored data commonly uses the partial log-likelihood
as a loss function
\citep[e.g.,][]{li2016regularized,yao2017deep,lee2018deephit,katzman2018deepsurv,gensheimer2019scalable,lee2021boosted,kvamme2021continuous}.
This loss function is also suggested for super learning with
right-censored data by \cite{polley2011-sl-cens}, where data
is assumed to be observed in discrete time. However, the
partial log-likelihood loss does not work well as a measure of
performance in hold-out samples observed in continuous time.
The reason is that the partial log-likelihood loss assigns an
infinite value for any learner that predicts piece-wise
constant cumulative hazard functions when there are event
times in the test set which do not occur in the learning set.
This problem occurs with prominent survival learners including
the Kaplan-Meier estimator, random survival forests, and
semi-parametric Cox regression models, and these learners
cannot be included in the library of the super learner
proposed by \cite{polley2011-sl-cens}. When a proportional
hazards model is assumed, the baseline hazard function can be
profiled out of the likelihood \citep{cox1972regression}. The
cross-validated partial log-likelihood loss
\citep{verweij1993cross} has therefore been suggested as a
loss function for super learning by \cite{golmakani2020super}.
This choice of loss function restricts the library of learners
to include only Cox proportional hazards models, and hence
excludes many learners such as, e.g., random survival forests,
additive hazards models, and accelerated failure time models.

@@

More recent work by \cite{han2021inverse} and
\cite{westling2021inference} circumvents the need to
pre-specify a censoring model by iterating between estimation
of the outcome and censoring models. However, this procedure
is in general not guaranteed to converge to the true
data-generating mechanism
\citep[][Appendix~A.4]{munch2024thesis}.

@@

We introduce our notation and framework in
Section~\ref{sec:framework}. Section~\ref{sec:super-learning}
defines general super learning for right-censored data. In
Section~\ref{sec:super-learner-simple} we define the state
learner, while Section~\ref{sec:theor-results-prop} provides
theoretical guarantees. In Section~\ref{sec:use-cases-state}
we demonstrate how the state learner can be used to obtain
risk predictions, and we briefly mention how it can be applied
in the context of causal inference.
Section~\ref{sec:numer-exper} reports the results of numerical
experiments, and Section~\ref{sec:real-data-appl} illustrates
the method on prostate cancer data. We conclude with a
discussion in Section~\ref{sec:discussion}. Proofs are
collected in Appendix~\ref{sec:proofs}. Code and an
implementation of the state learner are available at
\url{https://github.com/amnudn/statelearner}.

@@

\begin{proposition}
  \label{prop:stric-prop}
  If \(P \in\mathcal{P}\) then
  \begin{equation*}
    F_P = \argmin_{F \in \mathcal{H}_{\mathcal{P}}} P{[\bar{B}_\tau(F, \blank)]},
  \end{equation*}
  for all \( j \in \{-1, 0, 1, 2 \} \), almost all
  \( t \in [0,\tau] \), and \( P \)-almost all
  \( x \in \mathcal{X} \).
\end{proposition}
\begin{proof}
  See Appendix~\ref{sec:consistency-proof}.
\end{proof}

@@

Proposition~\ref{prop:oracle-prop} provides a bound on the
average price we pay for using cross-validation to select a
learner from a library of candidates as compared to knowing
the oracle learner in advance. Corollary~\ref{cor:asymp-cons}
states that this price is asymptotically vanishing, up to
logarithmic terms, as long as the number of learners in the
library grows with sample size at a polynomial rate.

@@

\subsection{Targeted machine learning and causal inference}
\label{sec:targeted-learning}

The framework of \textit{targeted} or \textit{debiased}
machine learning is a general methodology for combining
data-adaptive modeling, such as machine and super learning,
with asymptotically valid statistical inference for
interpretable target parameters \citep{van2006targeted,
  chernozhukov2018double}. \fxnote*{add more references?}{This
  framework has been applied extensively for causal inference
  based on observational data
  \citep[e.g.,][]{van2011targeted,kennedy2016semiparametric,rytgaard2022targeted}.}
A targeted or debiased estimator is constructed by first
estimation high-dimensional nuisance parameters using machine
learning, and then employing a debiasing or targeting step
that is motivated from semi-parametric efficiency theory
\citep{pfanzagl1985contributions,bickel1993efficient,van2003unified,tsiatis2007semiparametric,kennedy2022semiparametric}.
In the context of right-censored competing risk data, the
high-dimensional nuisance parameters needed to be estimated
will typically include the cause-specific cumulative hazard
functions and the cumulative hazard function for censoring
\citep[e.g.,][]{van2003unified,rytgaard2021estimation,rytgaard2022targeted}.
The state learner immediately provides estimates of these
nuisance parameters and is hence particularly well suited for
targeted and debiased machine learning. Though it served as
our original motivation for developing the state learner, we
leave the study of the state learner in the context of
targeted and debiased machine learning for a future paper.
